Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-09-27 document/ms.tex
Include content from GJH from overleaf

Starspots have recently been observed with exceptional photometric precision from monitoring surveys targeting exoplanet transits \citep[\emph{e.g}][]{harrison11,davenport15}.  The single band light curves from such surveys cannot separate the relative areas and temperatures of the emitting regions, and therefore cannot yield estimates for the filling factor of spots without making assumptions about the relative temperature contrast. Contemporaneous or near-contemporaneous panchromatic photometric monitoring \citep{petrov94,bouvier95,grankin07,cody14} can break the degeneracy between starspot filling factor and starspot temperature but these measurements are still limited to detecting large longitudinally asymmetric starspots.  Transiting planets or eclipsing binaries passing in front of starspots \citep{desert11} can break geometric degeneracies, providing estimates for the characteristic sizes and lifetimes of starspots or starspot groups, but the number of such transiting systems and demands of high photometric precision limit wider application.  Zeeman Doppler Imaging \citep[ZDI]{donati14} produces the latitudinal and longitudinal distribution of starspots in reconstructed brightness maps based on models for spectro-polarimetric line shapes, but would be unable to identify uniform distributions of small spots.  All of these techniques would tend to underestimate the starspot areal coverage of pole-on stars, stars with longitudinally symmetric bands of starspots, or stars with an isotropic distribution of small starspots.

@@

Rotational modulation is expected to lead to much smaller brightness changes at near-IR wavelengths than at optical wavelengths (see Table~\ref{tab:photrange}).  The smaller amplitude of near-IR brightness is also seen in a few spotted stars in the optical/IR monitoring of NGC 2264 \citep{cody14}, although those stars have a much smaller $V$ band amplitude than LkCa 4.

@@

\begin{table}
\caption{Predicted Photometric Range}
\centering
\label{tab:photrange}
\begin{tabular}{cccc}
\hline
Band & Low Spot & High Spot & Range\\
Cool Spot & 68\% & 83\% &\\
\hline
\multicolumn{4}{c}{Based on $V$, need to adapt for $A_V$}\\
$B$ & 13.80 & 14.48 & 0.68\\
$V$ & 12.56 & 13.23 & 0.67\\
$R$ & 11.53 & 12.17 & 0.64 \\
$I$ & 10.80 & 11.38 & 0.58\\
$J$ & 9.79 & 10.00 & 0.21\\
$K$ & 8.89 & 9.04 & 0.15\\
\hline
\end{tabular}
\end{table}

@@

%\subsection{Absolute lower limit of starspot filling factor}

@@

This simple thought experiment demonstrates that the high starspot filling factor must be approximately correct.  A lower limit to the starspot filling factor can be set by the size of a black spot that can induce such large $\Delta V$.  Consider the following extremely conservative assumptions:  \begin{enumerate}
  \item The starspot emits \emph{zero flux}.  This statement is equivalent to the starspot being at \emph{absolute zero}.
  \item The stellar disk shows no starspots at maximum brightness.
  \item The starspot transits the center of the stellar disk.
\end{enumerate}

For LkCa 4 in 2015 the $\Delta V\sim 0.5$ mag amplitude corresponds to an absolute minimum fill factor $f>0.37$.  At least 37\% of the stellar disk is covered in starspots at minimum light.  In the 2004 season, LkCa 4 exhibited $\Delta V\sim 0.8$, corresponding to a minimum fill factor of $f=0.52$.  If \emph{any} of these assumptions is relaxed, the minimum fill factor \emph{increases}.

@@

One conceivable interpretation of the 74\% to 86\% coverage fraction of cool photosphere is a stellar axis inclined towards our observing direction, with a large circum-polar spot covering much of the observable hemisphere, as described by \citet{donati14}.  The ambient hot photosphere would then be close to the equator, which would be limb-darkened due to projection effects.  Such limb darkening would cause a non-linear transformation between our observed fill factor $f_{\Omega}$, and the surface areal coverage fraction of spots.  It is conceivable, and probably likely, that the value we report for $f_{\Omega}$ is an overestimate of the surface area actually covered by spots.  In this circum-polar spot scenario, local fingers of cool spots could reach down to near the equator, entering into and out-of the field of view and giving rise to the large amplitude photometric variation seen on LkCa 4.

@@

We assumed that the \PHOENIX\ synthetic spectra are a good representation of the observed IGRINS spectrum of LkCa 4, a young star that possesses a strong magnetic field \citep{donati14}.  Cool sunspots are associated with locally heightened magnetic fields on the sun, so synthetic spectral models employing negligible magnetic fields will fail to fit magnetic sensitive features.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into the photosphere in the line wings than in the line center.  The former effect will preferentially impact lines with high magnetic sensitivity \citep[\emph{e.g.}][]{johnskrull99,deen13}.  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants \citep{oneal96}.  Our tacit assumptions have been that these line-broadening effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures, 2) there exist many orders and lines that are not affected at all, and 3) the Gaussian process noise model fosters some resilience against unknown residual spectrum correlations induced by such under-fitting.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$, as the na\"{i}ve fitting strategy futilely attempts to minimize the residuals from Zeeman-broadened lines.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.

@@

Radial velocity (RV) planet searches in the infrared--where the modulation of RV signals is lessened \citep[\emph{e.g.}][]{prato08}-- could potentially reduce RV jitter by characterizing and constraining the effect of direct emission from starspots, in addition to the flux deficits imbued in optical line profiles.  Direct spectral emission from cool starspots is likely much weaker in sources with less starspot coverage than LkCa 4.  As demands on RV precision increase, the characterization and modeling of direct emission from starspots should in principle decrease RV jitter attributable to starspots in the near-IR spectra of any spotted star.

@@

We thank Chris Kochanek for valuable comments on the draft.  MG-S and GJH also thank Steve Saar and GJH thanks Lynne Hillenbrand for valuable discussions about spots.

>>>>

Starspots have recently been observed with exceptional photometric precision from monitoring surveys targeting exoplanet transits \citep[\emph{e.g}][]{harrison11,davenport15}.  The single band light curves from such surveys cannot separate the relative areas and temperatures of the emitting regions, and therefore cannot yield estimates for the filling factor of spots without making assumptions about the relative temperature contrast. Contemporaneous or near-contemporaneous panchromatic photometric monitoring \citep{petrov94,bouvier95,grankin07,cody14} can break the degeneracy between starspot filling factor and starspot temperature but these measurements are still limited to detecting large longitudinally asymmetric starspots.  Transiting planets or eclipsing binaries passing in front of starspots \citep{desert11} can break geometric degeneracies, providing estimates for the characteristic sizes and lifetimes of starspots or starspot groups, but the number of such transiting systems and demands of high photometric precision limit wider application.  Zeeman Doppler Imaging \citep[ZDI]{donati14} produces the latitudinal and longitudinal distribution of starspots in reconstructed brightness maps based on models for spectro-polarimetric line shapes, but would be unable to identify uniform distributions of small spots.  All of these techniques would tend to underestimate the starspot areal coverage on stars with an isotropic distribution of small starspots, or pole-on stars or stars with longitudinally symmetric bands of starspots.

@@

Rotational modulation is expected to lead to much smaller brightness changes at near-IR wavelengths, with amplitudes of 0.15--0.2 mag, than the 0.6--0.8 mag amplitudes seen at optical wavelengths.  The smaller amplitude of near-IR brightness is also seen in a few spotted stars in the optical/IR monitoring of NGC 2264 \citep{cody14}, although those stars have a much smaller $V$ band amplitude than LkCa 4.

@@

%\begin{table}
%\caption{Predicted Photometric Range}
%\centering
%\label{tab:photrange}
%\begin{tabular}{cccc}
%\hline
%Band & Low Spot & High Spot & Range\\
%Cool Spot & 68\% & 83\% &\\
%\hline
%\multicolumn{4}{c}{Based on $V$, need to adapt for $A_V$}\\
%$B$ & 13.80 & 14.48 & 0.68\\
%$V$ & 12.56 & 13.23 & 0.67\\
%$R$ & 11.53 & 12.17 & 0.64 \\
%$I$ & 10.80 & 11.38 & 0.58\\
%$J$ & 9.79 & 10.00 & 0.21\\
%$K$ & 8.89 & 9.04 & 0.15\\
%\hline
%\end{tabular}
%\end{table}

@@

@@

This simple thought experiment demonstrates that the high starspot filling factor must be approximately correct.  A lower limit to the starspot filling factor can be set by the size of a black spot that can induce such large $\Delta V$.  Consider the following extremely conservative assumptions:  \begin{enumerate}
  \item The starspot emits \emph{zero flux}.  This statement is equivalent to the starspot being at \emph{absolute zero}.
  \item The stellar disk shows no starspots at maximum brightness.
  \item The starspot transits the center of the stellar disk.
  \item The star shows no feature brighter than the quiet photosphere at the visible surface
\end{enumerate}

For LkCa 4 in 2015 the $\Delta V\sim 0.5$ mag amplitude corresponds to an absolute minimum fill factor $f>0.37$.  At least 37\% of the stellar disk is covered in starspots at minimum light.  In the 2004 season, LkCa 4 exhibited $\Delta V\sim 0.8$, corresponding to a minimum fill factor of $f=0.52$.  If any of assumptions 1-3 is relaxed, the minimum fill factor \emph{increases}.  If assumption 4 is relaxed, the minimum fill factor decreases.  The large amplitude photometric variations in LkCa 4 are therefore evidence for either a hot spot in excess of the ambient photosphere, or a large coverage fraction of cool starspots.

@@

One conceivable interpretation of the 74\% to 86\% coverage fraction of cool photosphere is a stellar axis inclined towards our observing direction, with a large circum-polar spot covering much of the observable hemisphere, as described by \citet{donati14}.  The ambient hot photosphere would then be close to the equator, which would be limb-darkened due to projection effects.  Such limb darkening would cause a non-linear transformation between our observed fill factor $f_{\Omega}$, and the surface areal coverage fraction of spots.  It is conceivable, and probably likely, that the value we report for $f_{\Omega}$ is an overestimate of the surface area actually covered by spots.  In this circum-polar spot scenario, local fingers of cool spots could reach down to near the equator, entering into and out-of the field of view and giving rise to the large amplitude photometric variation seen on LkCa 4.  This scenario would have to be consistent with the inclination derived from relating $\vsini$, rotation period, and estimates for the stellar radius.  Such estimates are limited by estimates of the radius, since uncertainties in effective temperature and luminosity cause a wide range of radius estimates.

@@

We assumed that the \PHOENIX\ synthetic spectra are a good representation of the observed IGRINS spectrum of LkCa 4, a young star that possesses a strong magnetic field \citep{donati14}.  Cool sunspots are associated with locally heightened magnetic fields on the sun, so synthetic spectral models employing negligible magnetic fields will fail to fit magnetic sensitive features.  Magnetic fields will have a few effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into the photosphere in the line wings than in the line center.  The former effect will preferentially impact lines with high magnetic sensitivity \citep[\emph{e.g.}][]{johnskrull99,deen13}.  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants \citep{oneal96}.  Zeeman amplification can also be present \citep{basri92}. Our tacit assumptions have been that these line-broadening effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures, 2) there exist many orders and lines that are not affected at all, and 3) the Gaussian process noise model fosters some resilience against unknown residual spectrum correlations induced by such under-fitting.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$, as the na\"{i}ve fitting strategy futilely attempts to minimize the residuals from Zeeman-broadened lines.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.

@@

Radial velocity (RV) planet searches in the infrared--where the modulation of RV signals is lessened \citep[\emph{e.g.}][]{prato08,crockett12}-- could potentially reduce RV jitter by characterizing and constraining the effect of direct emission from starspots, in addition to the flux deficits imbued in optical line profiles.  Direct spectral emission from cool starspots is likely much weaker in sources with less starspot coverage than LkCa 4.  As demands on RV precision increase, the characterization and modeling of direct emission from starspots should in principle decrease RV jitter attributable to starspots in the near-IR spectra of any spotted star.

@@

We thank Chris Kochanek and Daniel T. Jaffe for valuable comments on the draft.  MG-S and GJH also thank Steve Saar and GJH thanks Lynne Hillenbrand for valuable discussions about spots.
